
```{r}
devtools::install_github("mikeyEcology/MLWIC2")
library(MLWIC2)
library(PressPurt)
options(timeout=600)

MLWIC2::setup(python_loc = "/Users/hayleeoyler/miniconda3/envs/r-reticulate/bin/python",gpu = FALSE, r_reticulate = TRUE)
MLWIC2:: setup()
```
Error message 1
WARNING conda.core.path_actions:verify(962): Unable to create environments file. Path not writable.
  #environment location: /Users/hayleeoyler/.conda/environments.txt

Error message 2
WARNING conda.core.envs_manager:register_env(50): Unable to register environment. Path not writable or missing.
  #environment location: /opt/anaconda3/envs/r-reticulate
  #registry file: /Users/hayleeoyler/.conda/environments.txt

/Users/hayleeoyler/Library/r-miniconda
python location: /Users/hayleeoyler/miniconda3/envs/r-reticulate/bin/python
conda location: /Users/hayleeoyler/miniconda3/condabin/conda 

```{r}
create_conda_env(condaenv="PressPurtEnv", 
                  version="/opt/anaconda3/bin/python3", 
                  verbose = TRUE)
py_depend(condaenv = "PressPurtEnv")
set_python_conda(condaenv="PressPurtEnv")
MLWIC2::runShiny('make_input')
setwd("/Users/hayleeoyler/Desktop/MLWIC_examples-master")

classify(path_prefix = "/Users/hayleeoyler/Desktop/MLWIC_examples-master/images", # this is the absolute path to the images. 
         data_info = "/Users/hayleeoyler/Desktop/MLWIC_examples-master/image_labels.csv", # this is the location of the csv containing image information. It has Unix linebreaks and no headers.
         model_dir = "/Users/hayleeoyler/Desktop/MLWIC_examples-master", # assuming this is where you stored the L1 folder in Step 3 of the instructions: github.com/mikeyEcology/MLWIC/blob/master/README
         python_loc = "/Users/hayleeoyler/miniconda3/envs/r-reticulate/bin/", # the location of Python on your computer. 
         save_predictions = "model_predictions.txt", # this is the default and you should use it unless you have reason otherwise.
         os = "Mac"
         )
```
Your `data_info` file exists: /Users/hayleeoyler/Desktop/MLWIC_examples-master/image_labels.csv.
Your `path_prefix exists: /Users/hayleeoyler/Desktop/MLWIC_examples-master/images.
You are not using a Windows computer.
/opt/anaconda3/bin/python: can't open file 'run.py': [Errno 2] No such file or directory
The classify function did not run properly.

```{r}
find_python()
#environment location: /Users/hayleeoyler/.conda/environments.txt

input_file(data_info.csv)
make_input(data_info.csv)

library(devtools) # load package
devtools::install_github("r-lib/pkgbuild") # install updated version of pkgbuild from GitHub
library(pkgbuild) # load package
find_rtools() # should be TRUE, assuming you have Rtools 3.5
```

