---
title: "R cheatsheet"
output: html_document
date: "2023-04-26"
---

```{r load-lake-data, include=FALSE}
library(readr)
lake_data <- read_csv("/Users/Haylee/myrepo/L05/R02_Data_mercury_lake.csv")
```

Note:  You can suppress the code and or output from r code by adding `message=FALSE` and/or `echo=False` to the code chunk description in brackets

## Reading in the data
- How to read a csv file using `read_csv()` 
- Get to know the data using `str()`, `head()`, `dim()`, and `names()` 
-  Manipulate the data frame using `dplyr`'s `rename()`, `select()`, `arrange()`,
`filter()`, `mutate()`, `group_by()`, and `summarize()

```{r initial-looks}
head(lake_data)
dim(lake_data)
names(lake_data)
str(lake_data)
```

head: shows first six rows of data
dim: gives dimensions of data
names: names of columns, names of all the variables in the data set
str: structure of the data, gives data frame, name of variables, the numeric/character, and some of the values. A general summary 

## Using `dplyr` functions to work with the data frame

## Function 1: rename()
```{r rename, echo=FALSE}
library(dplyr)

names(lake_data)
lake_data_tidy <- lake_data %>% rename(lake_name = lakes)
names(lake_data_tidy)

#You can rename multiple variables at once
lake_data_tidy <- lake_data %>% rename(lake_name = lakes, 
                                       ph_level = ph)
```

## Function 2: select()
```{r select}
smaller_data <- lake_data %>% select(lakes, ph, chlorophyll)
names(smaller_data)
#We use `select()` to select a subset of **variables**. 
#This is very handy if we inherit a large dataset with several variables that we do not need.

#We can also use "negative `select()`" to deselect variables. Suppose we wanted to keep all variables except for `age_data`
smaller_data_2 <- lake_data %>% select(- age_data)
names(smaller_data_2)
```

## Function 3: mutate()
- **`mutate()` is a very useful function**. 
- It is used to add new variables to the dataset. Suppose that someone told you that the number of fish sampled was actually in hundreds, such that 5 is actually 500. 

You can use mutate to add this new variable to your dataset
```{r mutate}
lake_data_new_fish <- lake_data %>% 
  mutate(actual_fish_sampled = number_fish_sampled * 100)
```

## Use `%>%` to "pipe" several lines of code together
You can think of the "pipe" operator as a "then" when you read code 

Translating the code below to words:

Tidy lake data is equal to lake data with a renamed lake variable, a mutate fish sample variable, and without age data and # fish sampled variables.
```{r pipe practice}
tidy_lake_data <- lake_data %>% 
  rename(lake_name = lakes) %>%
  mutate(actual_fish_sampled = number_fish_sampled * 100) %>%
  select(- age_data, - number_fish_sampled)

tidy_lake_data
```

## Function 4: `filter()`
```{r filter practice}
#Greater than, less than, greater/less than equal to. >=, <=, <, >
lake_data_filtered <- lake_data %>% filter(ph > 7)

#Equal to
lake_data %>% filter(age_data == "recent")

#Is not equal to
lake_data %>% filter(age_data != "recent") 

#In
lake_data %>% filter(lakes %in% c("Alligator", "Blue Cypress")) 

#Include multiple statements. The comma combines the filter statements, so we are filtering to including data with both `ph > 6` and `chlorophyll > 30`
lake_data %>% filter(ph > 6, chlorophyll > 30)

#Or, At least one of `ph > 6` or `chlorophyll > 30` needs to be true.
lake_data %>% filter(ph > 6 | chlorophyll > 30)
```

## Functions 5 and 6: group_by() and summarize()
`group_by() and summarize()`  group the data by a categorial variable and calculate a statistic.

`mean()` and `sd()` to calculate the mean and standard deviation of variables.
```{r group by and summarize}
lake_data %>% 
  group_by(age_data) %>% 
  summarize(mean_ph = mean(ph),
            standard_deviation_ph = sd(ph))
```


## ggplot
## `geom_` commands so far
- `geom_histogram()` to make histograms
- `geom_point()` to make scatter plots
- `geom_line()` to make line plots
- `geom_boxplot()` to make box plots
- `geom_bar() to make a bar graph

## some useful options for ggplots
- `col = "blue"`, `size = 2`, `lty = 2`, to change color, size and line type of the `geom`
- `aes(col = race)` to *link* color to race
- `aes(lty = sex)` to *link* line type to sex
- `facet_wrap(~ var1)` to make separate plots for different levels of one variable
- `facet_grid(var1 ~ var2)` to make separate plots for combinations of levels of two variables
- `stat=identity` vs `stat=count` to tell ggplot whether it needs to calculate the percents
- `aes(fill=sex)` to *link* color to sex in a histogram

Say we want to make a scatter plot of the life expectancy for white males in California over time.
```{r ggplot-canvas, fig.width=4, fig.height=3, echo=FALSE}
le_data <- read_csv("Data/Life-expectancy-by-state-long.csv")
wm_cali <- le_data %>% filter(state == "California", sex == "Male", 
                              race == "white")

## First step to building a `ggplot()`: set up the canvas
library(ggplot2)
ggplot(data = wm_cali, aes(x = year, y = LE)) + theme_minimal(base_size = 15)

## Second step to building a `ggplot()`: tell `ggplot` how to plot the data
ggplot(data = wm_cali, aes(x = year, y = LE)) + geom_point()

## `labs()` to add a title, a caption, and modify x and y axes titles
## `col` controls the color of geom_point()
## `size` controls the size of geom_point()
ggplot(data = wm_cali, aes(x = year, y = LE)) + geom_point(col = "blue", size = 4) +
  labs(title = "Life expectancy in white men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
```

## Life expectancy for white and black men in California
What do we need to change to make a separate line for both black and white men?
```{r plots}
## First, update the `filter()`
wbm_cali <- le_data %>% filter(state == "California",
                               sex == "Male")

## And change previous graph to **link** color to race
ggplot(data = wbm_cali, aes(x = year, y = LE)) + geom_line(aes(col = race)) +
  labs(title = "Life expectancy in black and white men in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")

```

## What if we also wanted to look at women?
```{r more plot permutations}
cali_data <- le_data %>% filter(state == "California")

## What is wrong with this plot?
ggplot(data = cali_data, aes(x = year, y = LE)) + geom_line(aes(col = race)) +
  labs(title = "Life expectancy in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")
#This just colored by race group and it connected all the dots, there's space btwn male and female operations which made it all squiggly. 

## Use `lty()` to **link** line type to sex
ggplot(data = cali_data, aes(x = year, y = LE)) + geom_line(aes(col = race, lty = sex)) +
  labs(title = "Life expectancy in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)")

## Use `facet_wrap()` to make separate plots for a specified variable
ggplot(data = cali_data, aes(x = year, y = LE)) + 
  geom_line(aes(col = race, lty = sex)) +
  labs(title = "Life expectancy in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)") +
  facet_wrap(~ sex)

## What would this look like as a bar chart?
ggplot(data = cali_data, aes(x =year, y=LE)) + 
  geom_line(stat = "identity", aes(col=race)) +
  labs(title = "Life expectancy in California, 1969-2013",
       y = "Life expectancy", 
       x = "Year", 
       caption = "Data from Riddell et al. (2018)") +
  facet_wrap(~ sex + race)

## Draw a histogram for Life expectancy in one year (2013)
data_2013 <- le_data %>% filter(year == 2013)

ggplot(dat = data_2013, aes(x = LE)) + 
  geom_histogram(binwidth = 1, col = "white", aes(fill = sex))  +
    facet_grid(race ~ sex)
#facet_grid also allows for multiple plots

```


# Regression
## What is a regression line?
- A straight line that is **fitted** to data to minimize the distance between the data and the fitted line. 
- It is often called the **line of best fit**. 
- The line of best fit can be used to describe the relationship between the 
explanatory and response variables.
- Can you fit a line of best fit for non-linear relationships?
- Very important to visualize the relationship first. Why?

## Equation of the line of best fit
The line of best fit can be represented by the equation for a line:
a= intercept, b=slope

$$y = a + bx$$

## Fitting a linear model in R

- `lm()` is the function for a linear model.
- The first argument that `lm()` wants is a formula `y ~ x`.
    - `y` is the response variable from your dataset
    - `x` is the explanatory variable
- The second argument sent to `lm()` is the data set.
```{r lm template}
lm(formula = y ~ x, data = your_dataset)

#Recall the manatee example from Ch.3 that examined the relationship between the number of registered `powerboats` and the number of manatee `deaths` in Florida between 1977 and 2016
mana_data <- read_csv("Data/Ch03_Manatee-deaths.csv")

#Recall that the relationship was linear by examining the scatter plot:
mana_death1<-ggplot(mana_data, aes(x = powerboats, y = deaths)) +
  geom_point() +
  theme_minimal(base_size = 15)

#Calculate the line of best fit: 
mana_lm <- lm(deaths ~ powerboats, mana_data)
library(broom) # we use the package broom to look at the output of the linear model
tidy(mana_lm) # only pay attention to the term and estimate columns for now.

# Add the regression line to the scatter plot using `geom_abline()`
default_zoom_plot <- ggplot(mana_data, aes(x = powerboats, y = deaths)) +
  geom_point() + 
  labs(x = "Powerboats registered (X 1000)",
       y = "Manatee deaths") +
  geom_abline(intercept = -46.7520, slope = 0.1358) + 
  coord_fixed(ratio = 5) +
  theme_minimal(base_size = 15)
```
- Intercept: The predicted number of deaths if there were no powerboats. But the prediction is negative. Why?
- Powerboats: This is the slope. What does the estimate slope for powerboats mean?

## R-squared

When we run a linear model, the r-squared is also calculated. Here is how to see the r-squared for the manatee data:
```{r glance}
glance(mana_lm)
#Interpretation of r-squared: The fraction of the variation in the values of y that is explained by the line of best fit.
```


## Example using transformed data

- Sometimes, the data is transformed to another scale so that the relationship
between the transformed $x$ and $y$ is linear
- Table 3.4 in B&M provides data on the mean number of seeds produced in a year 
by several common tree species and the mean weight (im milligrams) of the seeds 
produced. 
```{r make seed dataset, echo=FALSE}
library(tibble)
seed_data <- tribble(~ species, ~ seed_count, ~ seed_weight,
                       "Paper birch", 27239, 0.6,
                       "Yellow birch", 12158, 1.6,
                       "White spruce", 7202, 2.0,
                       "Engelman spruce", 3671, 3.3, 
                       "Red spruce", 5051, 3.4, 
                       "Tulip tree", 13509, 9.1, 
                       "Ponderosa pine", 2667, 37.7, 
                       "White fir", 5196, 40.0, 
                       "Sugar maple", 1751, 48.0, 
                       "Sugar pine", 1159, 216.0, 
                       "American beech", 463, 247, 
                       "American beech", 1892, 247,
                       "Black oak", 93, 1851, 
                       "Scarlet oak", 525, 1930, 
                       "Red oak", 411, 2475, 
                       "Red oak", 253, 2475,
                       "Pignut hickory", 40, 3423, 
                       "White oak", 184, 3669, 
                       "Chestnut oak", 107, 4535)

## Scatter plot of `seed_weight` vs. `seed_count`
ggplot(seed_data, aes(seed_count, seed_weight)) + 
  geom_point() +
  theme_minimal(base_size = 15)
#- `seed_count` and `seed_weight` both vary widely. Their relationship is not linear
```

## Investigate the relationship between their logged variables

- Add transformed variables to the dataset using `mutate()`. 
- We add both log base $e$ and log base 10 variables for illustration
```{r plot and interpret}
seed_data <- seed_data %>% mutate(log_seed_count = log(seed_count), 
                                  log_seed_weight = log(seed_weight),
                                  log_b10_count = log(seed_count, 10),
                                  log_b10_weight = log(seed_weight, 10))

## Plot transformed data (log base e)
ggplot(seed_data, aes(log_seed_count, log_seed_weight)) + 
  geom_point() +
  labs(x = "Log of seed count", y = "Log of seed weight", 
       title = "Using the natural log (base e)") +
  theme_minimal(base_size = 15)

## Plot transformed data (log base 10)
ggplot(seed_data, aes(log_b10_count, log_b10_weight)) + 
  geom_point() +
  labs(x = "Log of seed count", y = "Log of seed weight", 
       title = "Using log base 10") +
  theme_minimal(base_size = 15)

#You can use either base 10 or base $e$ for class.
#The calculations using base $e$ are easier

## `lm()` on the log (base e) variables
seed_mod <- lm(log_seed_weight ~ log_seed_count, data = seed_data)
tidy(seed_mod)
glance(seed_mod) %>% pull(r.squared)

## `lm()` on the log (base 10) variables
seed_mod_b10 <- lm(log_b10_weight ~ log_b10_count, data = seed_data)
tidy(seed_mod_b10)
glance(seed_mod_b10) %>% pull(r.squared)
#What is different from the log base $e$ output?

```

## Interpretation of `lm()` when using log (base $e$) data

- We use the results of the `lm()` on the log (base $e$) transformed data for making 
predictions
- E.g., what seed weight is predicted for a seed count of 2000?
- Worked calculation:

1. Write down the line of best fit: $log_e(seed.weight) = 15.49130 - 1.522220\times{log_e(seed.count)}$
2. Plug in $seed.count = 2000$ into the line of best fit: $log_e(seed.weight) = 15.49130 - 1.522220\times{log_e(2000)}$
3. Solve for seed count by exponentiating both sides: 
$$seed.weight = exp(15.49130 - 1.522220\times{log_e(2000)})$$ (this uses the property that $e^{log_e(x)}=x$)
$$seed.weight = 50.45$$

4. Interpret: Seeds are expected to weigh 50.45 for trees having a seed count of 2000.